---
title: "neural network"
output: html_document
date: "2025-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Package Installation**
```{r}
library(MASS) 
library(dplyr)
library(nnet)
```

**Training Dataset Simulation**
```{r}
set.seed(814)

N <- 100   # number of individuals
T <- 10    # number of observations per individual
price_range <- c(3, 8)

#Simulate individual-level covariates
age <- sample(18:70, N, replace = TRUE)
gender <- sample(c(0, 1), N, replace = TRUE)        # 0 = female, 1 = male
urban <- sample(c(0, 1), N, replace = TRUE)         # 0 = rural, 1 = urban

# Create design matrix for individual-level covariates (excluding price)
indiv_covariates <- data.frame(id = 1:N, age = age, gender = gender, urban = urban)

#Simulate individual-specific coefficients: beta_i ~ N(mu_, Sigma_)
#intercept, price, age, gender, urban
mu_ <- c(0, -0.8, 0.01, 0.5, 0.3)
Sigma_ <- matrix(c(
  1.5, 0.1,  0.05, 0.1, 0.05,
  0.1, 0.5, -0.02, 0.0, 0.02,
  0.05, -0.02, 0.2, 0.01, 0.0,
  0.1,  0.0,  0.01, 0.4, 0.03,
  0.05, 0.02, 0.0,  0.03, 0.3
), nrow = 5, byrow = TRUE)

beta_i <- mvrnorm(n = N, mu = mu_, Sigma = Sigma_)

#Simulate data using hierarchical logit model
sim_binary <- data.frame()

for (i in 1:N) {
  for (t in 1:T) {
    price <- runif(1, price_range[1], price_range[2])
    X <- c(1, price, age[i], gender[i], urban[i])  # feature vector
    utility <- sum(X * beta_i[i, ])
    p <- 1 / (1 + exp(-utility))                   # logistic probability
    y <- rbinom(1, 1, p)                           # purchase decision

    sim_binary <- rbind(sim_binary, data.frame(
      id = i,
      time = t,
      price = price,
      age = age[i],
      gender = gender[i],
      urban = urban[i],
      purchase = y
    ))
  }
}

head(sim_binary)

```

**distribution of price coefficient**
```{r}
# Extract the price coefficients
price_coef <- beta_i[, 2]  

# Create a histogram
hist(price_coef, 
     breaks = 15, 
     col = "skyblue", 
     main = "",
     xlab = "Price Coefficient (β_price)",
     ylab = "Frequency")

lines(density(price_coef), col = "red", lwd = 2)
abline(v = mean(price_coef), col = "darkgreen", lwd = 2, lty = 2)
summary(price_coef)

```

**ground truth construction of training dataset**
```{r}
# Define purchase probability and expected revenue functions
purchase_prob <- function(X, beta) {
  plogis(sum(X * beta))
}

expected_revenue <- function(price, X, beta) {
  price * purchase_prob(X, beta)
}

price_options <- c(3, 8)
ground_truth <- data.frame()

for (i in 1:N) {
  indiv_covs <- c(age[i], gender[i], urban[i])
  X3 <- c(1, 3, indiv_covs)
  X8 <- c(1, 8, indiv_covs)
  
  prob_3 <- purchase_prob(X3, beta_i[i, ])
  prob_8 <- purchase_prob(X8, beta_i[i, ])
  
  rev_3 <- expected_revenue(3, X3, beta_i[i, ])
  rev_8 <- expected_revenue(8, X8, beta_i[i, ])
  
  # Determine the optimal price based on ground truth
  opt_price <- ifelse(rev_3 > rev_8, 3, 8)
  
  # Add both rows for each individual
  ground_truth <- rbind(
    ground_truth,
    data.frame(
      id = i,
      price = 3,
      predicted_prob = prob_3,
      predicted_revenue = rev_3,
      optimal_price = opt_price
    ),
    data.frame(
      id = i,
      price = 8,
      predicted_prob = prob_8,
      predicted_revenue = rev_8,
      optimal_price = opt_price
    )
  )
}

head(ground_truth)

total_optimal_revenue <- ground_truth %>%
  filter(price == optimal_price) %>%
  summarise(total_revenue = sum(predicted_revenue)) %>%
  pull(total_revenue)

cat("Total ground truth optimal revenue:", total_optimal_revenue, "\n")

```

**frequncy table of two prices under the ground truth pricing policy on training dataset**
```{r}
count <- ground_truth %>%
  group_by(optimal_price) %>%
  summarise(count = n() / 2) %>%
  arrange(optimal_price)

print(count)
```
**fitting Neural Network on training dataset**
```{r}

set.seed(814)

nn_model <- nnet(
  factor(purchase) ~ price + age + gender + urban,
  data = sim_binary,
  size = 5,          # hidden units
  maxit = 500,       # iterations
  decay = 1e-4,      # weight decay for regularization
  trace = FALSE      # suppress output
)

#Construct covariate matrix
individual_covs <- sim_binary %>%
  group_by(id) %>%
  dplyr::slice(1) %>%    
  ungroup() %>%
  select(id, age, gender, urban)


price_given <- c(3, 8)

nndata <- individual_covs %>%
  tidyr::crossing(price = price_given) %>%
  arrange(id, price)

# Predict purchase probabilities
nn_pred_probs <- predict(nn_model, newdata = nndata, type = "raw")

# Compute expected revenue
nndata <- nndata %>%
  mutate(
    nn_predicted_prob = as.numeric(nn_pred_probs),
    nn_predicted_revenue = nn_predicted_prob * price
  )

#Find optimal price under neural network policy
nn_best_prices <- nndata %>%
  group_by(id) %>%
  slice_max(nn_predicted_revenue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(id, nn_optimal_price = price)

nndata <- nndata %>%
  left_join(nn_best_prices, by = "id")
head(nndata)


```

**frequncy table of two prices under the neural network pricing policy on training dataset**
```{r}
nn_counts <- nndata %>%
  group_by(nn_optimal_price) %>%
  summarise(count = n() / 2) %>%
  arrange(nn_optimal_price)

print(nn_counts)
```


**Calculation of the expected total revenue based on the neural network pricing policy and the ground truth individual preferences**
```{r}

purchase_prob <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  linpred <- alpha + beta_price * p + beta_age * a + beta_gender * g + beta_urban * u
  plogis(linpred)
}

expected_revenue <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  p * purchase_prob(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban)
}

nn_eval_train <- nn_best_prices %>%
  left_join(individual_covs, by = "id") %>%
  mutate(
    alpha       = beta_i[id, 1],
    beta_price  = beta_i[id, 2],
    beta_age    = beta_i[id, 3],
    beta_gender = beta_i[id, 4],
    beta_urban  = beta_i[id, 5]
  )

nn_eval_train <- nn_eval_train %>%
  mutate(
    true_prob = purchase_prob(nn_optimal_price, age, gender, urban,
                              alpha, beta_price, beta_age, beta_gender, beta_urban),
    true_revenue = expected_revenue(nn_optimal_price, age, gender, urban,
                                    alpha, beta_price, beta_age, beta_gender, beta_urban)
  )

total_nn_policy_revenue_train <- sum(nn_eval_train$true_revenue)

cat("Total expected revenue on training data using NN policy:",
    total_nn_policy_revenue_train, "\n")

```


**hit rate on training dataset**
```{r}

nn_pred_prob_train <- predict(nn_model, newdata = sim_binary, type = "raw")

nn_pred_binary_train <- as.integer(nn_pred_prob_train > 0.5)

nn_validation_train_df <- sim_binary %>%
  mutate(
    nn_predicted_prob = as.numeric(nn_pred_prob_train),
    nn_predicted_purchase = nn_pred_binary_train
  ) %>%
  select(id, time, price, age, gender, urban, purchase,
         nn_predicted_prob, nn_predicted_purchase)

head(nn_validation_train_df) 

nn_hit_rate_train <- mean(nn_validation_train_df$nn_predicted_purchase == nn_validation_train_df$purchase)

cat("In-sample Hit Rate for Neural Network on Training Data:", nn_hit_rate_train, "\n")


```
**testing data simulation**
```{r}

set.seed(999) 
N_test <- 20
T <- 10
price_range <- c(3, 8)

age_test <- sample(18:70, N_test, replace = TRUE)
gender_test <- sample(c(0, 1), N_test, replace = TRUE)
urban_test <- sample(c(0, 1), N_test, replace = TRUE)

beta_i_test <- mvrnorm(n = N_test, mu = mu_, Sigma = Sigma_)

sim_binary_test <- data.frame()

for (i in 1:N_test) {
  for (t in 1:T) {
    price <- runif(1, price_range[1], price_range[2])
    X <- c(1, price, age_test[i], gender_test[i], urban_test[i])
    utility <- sum(X * beta_i_test[i, ])
    p <- 1 / (1 + exp(-utility))
    y <- rbinom(1, 1, p)
    
    sim_binary_test <- rbind(sim_binary_test, data.frame(
      id = i + N,  # ensure IDs don't overlap with training data
      time = t,
      price = price,
      age = age_test[i],
      gender = gender_test[i],
      urban = urban_test[i],
      purchase = y
    ))
  }
}

head(sim_binary_test)

```


**ground truth for the testing data**
```{r}
purchase_prob <- function(X, beta) {
  plogis(sum(X * beta))
}

expected_revenue <- function(price, X, beta) {
  price * purchase_prob(X, beta)
}
price_options <- c(3, 8)
ground_truth_test <- data.frame()

# Loop through each of the 20 out-of-sample individuals
for (i in 1:N_test) {
  # Extract covariates for test individuals
  indiv_covs <- c(age_test[i], gender_test[i], urban_test[i])

  X3 <- c(1, 3, indiv_covs)
  X8 <- c(1, 8, indiv_covs)
  
  prob_3 <- purchase_prob(X3, beta_i_test[i, ])
  prob_8 <- purchase_prob(X8, beta_i_test[i, ])
  
  rev_3 <- expected_revenue(3, X3, beta_i_test[i, ])
  rev_8 <- expected_revenue(8, X8, beta_i_test[i, ])
  
  opt_price <- ifelse(rev_3 > rev_8, 3, 8)
  
  ground_truth_test <- rbind(
    ground_truth_test,
    data.frame(
      id = i + N,   # IDs start after training data
      price = 3,
      predicted_prob = prob_3,
      predicted_revenue = rev_3,
      optimal_price = opt_price
    ),
    data.frame(
      id = i + N,
      price = 8,
      predicted_prob = prob_8,
      predicted_revenue = rev_8,
      optimal_price = opt_price
    )
  )
}

head(ground_truth_test)

total_optimal_revenue_test <- ground_truth_test %>%
  filter(price == optimal_price) %>%
  summarise(total_revenue = sum(predicted_revenue)) %>%
  pull(total_revenue)

cat("Total ground truth optimal revenue for TEST sample:", total_optimal_revenue_test, "\n")


```


**Fitting the trained neural network on testing data, find the pricing policy and the expected total revenue  **
```{r}

# Build testing covariates
testing_covs <- sim_binary_test %>%
  group_by(id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(id, age, gender, urban)

price_given <- c(3, 8)

nndata_test <- testing_covs %>%
  tidyr::crossing(price = price_given) %>%
  arrange(id, price)

# Predict purchase probabilities using the fitted nn_model
nn_pred_probs_test <- predict(nn_model, newdata = nndata_test, type = "raw")

nndata_test <- nndata_test %>%
  mutate(
    nn_predicted_prob = as.numeric(nn_pred_probs_test),
    nn_predicted_revenue = nn_predicted_prob * price
  )

nn_best_prices_test <- nndata_test %>%
  group_by(id) %>%
  slice_max(nn_predicted_revenue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(id, nn_optimal_price = price)

nndata_test <- nndata_test %>%
  left_join(nn_best_prices_test, by = "id")

head(nndata_test)

purchase_prob <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  linpred <- alpha + beta_price * p + beta_age * a + beta_gender * g + beta_urban * u
  plogis(linpred)
}

expected_revenue <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  p * purchase_prob(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban)
}

# Evaluate NN pricing policy on testing data
nn_eval_test <- nn_best_prices_test %>%
  left_join(testing_covs, by = "id") %>%
  mutate(
    alpha       = beta_i_test[id - N, 1],
    beta_price  = beta_i_test[id - N, 2],
    beta_age    = beta_i_test[id - N, 3],
    beta_gender = beta_i_test[id - N, 4],
    beta_urban  = beta_i_test[id - N, 5]
  )

# Define purchase probability & revenue using ground truth
nn_eval_test <- nn_eval_test %>%
  mutate(
    true_prob = purchase_prob(nn_optimal_price, age, gender, urban,
                              alpha, beta_price, beta_age, beta_gender, beta_urban),
    true_revenue = expected_revenue(nn_optimal_price, age, gender, urban,
                                    alpha, beta_price, beta_age, beta_gender, beta_urban)
  )

total_nn_policy_revenue_test <- sum(nn_eval_test$true_revenue)

cat("Total expected revenue on testing data using NN policy:",
    total_nn_policy_revenue_test, "\n")


```

**frequncy table of two prices under the Neural Network pricing policy on testing dataset**
```{r}
nn_counts2 <- nndata_test %>%
  group_by(nn_optimal_price) %>%
  summarise(count = n() / 2) %>%
  arrange(nn_optimal_price)

print(nn_counts2)
```
**hit rate on testing data**
```{r}
nn_pred_prob_test <- predict(nn_model, newdata = sim_binary_test, type = "raw")

# Convert predicted probabilities to binary purchase predictions
nn_pred_binary_test <- as.integer(nn_pred_prob_test > 0.5)

# Create validation dataframe for testing data
nn_validation_test_df <- sim_binary_test %>%
  mutate(
    nn_predicted_prob = as.numeric(nn_pred_prob_test),
    nn_predicted_purchase = nn_pred_binary_test
  ) %>%
  select(id, time, price, age, gender, urban, purchase,
         nn_predicted_prob, nn_predicted_purchase)

head(nn_validation_test_df)
# Calculate out-of-sample hit rate
nn_hit_rate_test <- mean(nn_validation_test_df$nn_predicted_purchase == nn_validation_test_df$purchase)

cat("Out-of-sample Hit Rate for Neural Network on Testing Data:", nn_hit_rate_test, "\n")
```






