---
title: "XGBoost"
output: html_document
date: "2025-08-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Package Installation**
```{r}
library(MASS) 
library(dplyr)
library(xgboost)
library(tidyr)
```

**Training Dataset Simulation**
```{r}
set.seed(814)

N <- 100   # number of individuals
T <- 10    # number of observations per individual
price_range <- c(3, 8)

#Simulate individual-level covariates
age <- sample(18:70, N, replace = TRUE)
gender <- sample(c(0, 1), N, replace = TRUE)        # 0 = female, 1 = male
urban <- sample(c(0, 1), N, replace = TRUE)         # 0 = rural, 1 = urban

# Create design matrix for individual-level covariates (excluding price)
indiv_covariates <- data.frame(id = 1:N, age = age, gender = gender, urban = urban)

#Simulate individual-specific coefficients: beta_i ~ N(mu_, Sigma_)
#intercept, price, age, gender, urban
mu_ <- c(0, -0.8, 0.01, 0.5, 0.3)
Sigma_ <- matrix(c(
  1.5, 0.1,  0.05, 0.1, 0.05,
  0.1, 0.5, -0.02, 0.0, 0.02,
  0.05, -0.02, 0.2, 0.01, 0.0,
  0.1,  0.0,  0.01, 0.4, 0.03,
  0.05, 0.02, 0.0,  0.03, 0.3
), nrow = 5, byrow = TRUE)

beta_i <- mvrnorm(n = N, mu = mu_, Sigma = Sigma_)

#Simulate data using hierarchical logit model
sim_binary <- data.frame()

for (i in 1:N) {
  for (t in 1:T) {
    price <- runif(1, price_range[1], price_range[2])
    X <- c(1, price, age[i], gender[i], urban[i])  # feature vector
    utility <- sum(X * beta_i[i, ])
    p <- 1 / (1 + exp(-utility))                   # logistic probability
    y <- rbinom(1, 1, p)                           # purchase decision

    sim_binary <- rbind(sim_binary, data.frame(
      id = i,
      time = t,
      price = price,
      age = age[i],
      gender = gender[i],
      urban = urban[i],
      purchase = y
    ))
  }
}

head(sim_binary)

```

**distribution of price coefficient**
```{r}
# Extract the price coefficients
price_coef <- beta_i[, 2]  

# Create a histogram
hist(price_coef, 
     breaks = 15, 
     col = "skyblue", 
     main = "",
     xlab = "Price Coefficient (β_price)",
     ylab = "Frequency")

lines(density(price_coef), col = "red", lwd = 2)
abline(v = mean(price_coef), col = "darkgreen", lwd = 2, lty = 2)
summary(price_coef)

```


**ground truth construction of training dataset**
```{r}
# Define purchase probability and expected revenue functions
purchase_prob <- function(X, beta) {
  plogis(sum(X * beta))
}

expected_revenue <- function(price, X, beta) {
  price * purchase_prob(X, beta)
}

price_options <- c(3, 8)
ground_truth <- data.frame()

for (i in 1:N) {
  indiv_covs <- c(age[i], gender[i], urban[i])
  X3 <- c(1, 3, indiv_covs)
  X8 <- c(1, 8, indiv_covs)
  
  prob_3 <- purchase_prob(X3, beta_i[i, ])
  prob_8 <- purchase_prob(X8, beta_i[i, ])
  
  rev_3 <- expected_revenue(3, X3, beta_i[i, ])
  rev_8 <- expected_revenue(8, X8, beta_i[i, ])
  
  # Determine the optimal price based on ground truth
  opt_price <- ifelse(rev_3 > rev_8, 3, 8)
  
  # Add both rows for each individual
  ground_truth <- rbind(
    ground_truth,
    data.frame(
      id = i,
      price = 3,
      predicted_prob = prob_3,
      predicted_revenue = rev_3,
      optimal_price = opt_price
    ),
    data.frame(
      id = i,
      price = 8,
      predicted_prob = prob_8,
      predicted_revenue = rev_8,
      optimal_price = opt_price
    )
  )
}

head(ground_truth)

total_optimal_revenue <- ground_truth %>%
  filter(price == optimal_price) %>%
  summarise(total_revenue = sum(predicted_revenue)) %>%
  pull(total_revenue)

cat("Total ground truth optimal revenue:", total_optimal_revenue, "\n")

```

**frequncy table of two prices under the ground truth pricing policy on training dataset**
```{r}
count <- ground_truth %>%
  group_by(optimal_price) %>%
  summarise(count = n() / 2) %>%
  arrange(optimal_price)

print(count)
```

**fitting XGBoost on training dataset**
```{r}

xgb_train <- sim_binary %>%
  select(price, age, gender, urban) %>%
  as.matrix()

y_train <- sim_binary$purchase

#fit XGBoost on the entire training dataset
set.seed(814)
xgb_model <- xgboost(
  data = xgb_train,
  label = y_train,
  objective = "binary:logistic",  # For classification
  nrounds = 100,
  verbose = 0
)

price_options <- c(3, 8)

individual_covs <- sim_binary %>%
  group_by(id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  select(id, age, gender, urban)

xgb_newdata <- individual_covs %>%
  crossing(price = price_options) %>%
  arrange(id, price)

xgb_matrix <- xgb_newdata %>%
  select(price, age, gender, urban) %>%
  as.matrix()

#Predict purchase probabilities
xgb_pred_probs <- predict(xgb_model, newdata = xgb_matrix)

#Compute predicted revenue
xgb_predictions <- xgb_newdata %>%
  mutate(
    xgb_predicted_prob = xgb_pred_probs,
    xgb_predicted_revenue = xgb_predicted_prob * price
  )

#Optimal price per individual
xgb_best_prices <- xgb_predictions %>%
  group_by(id) %>%
  slice_max(xgb_predicted_revenue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(id, xgb_optimal_price = price)

xgb_predictions <- xgb_predictions %>%
  left_join(xgb_best_prices, by = "id")

head(xgb_predictions)

```
**frequncy table of two prices under the XGBoost pricing policy on training dataset**
```{r}
xgb_counts <- xgb_predictions %>%
  group_by(xgb_optimal_price) %>%
  summarise(count = n() / 2) %>%
  arrange(xgb_optimal_price)

print(xgb_counts)
```


**Calculation of the expected total revenue based on the XGBoost pricing policy and the ground truth individual preferences**
```{r}
# Define purchase probability and expected revenue functions using scalar inputs
purchase_prob <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  linpred <- alpha + beta_price * p + beta_age * a + beta_gender * g + beta_urban * u
  plogis(linpred)
}

expected_revenue <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  p * purchase_prob(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban)
}

# Join XGBoost optimal prices with individual-level covariates and true coefficients
xgb_eval <- xgb_best_prices %>%
  left_join(individual_covs, by = "id") %>%
  mutate(
    alpha        = beta_i[id, 1],
    beta_price   = beta_i[id, 2],
    beta_age     = beta_i[id, 3],
    beta_gender  = beta_i[id, 4],
    beta_urban   = beta_i[id, 5]
  )


xgb_eval <- xgb_eval %>%
  mutate(
    xgb_true_prob = purchase_prob(xgb_optimal_price, age, gender, urban,
                                  alpha, beta_price, beta_age, beta_gender, beta_urban),
    
    xgb_true_revenue = expected_revenue(xgb_optimal_price, age, gender, urban,
                                        alpha, beta_price, beta_age, beta_gender, beta_urban)
  )

total_xgb_policy_revenue <- sum(xgb_eval$xgb_true_revenue)

cat("Total revenue using XGBoost policy (evaluated with true preferences):",
    total_xgb_policy_revenue, "\n")


```


**hit rate on training dataset**
```{r}

#Prepare design matrix using actual prices in sim_binary
xgb_insample_matrix <- sim_binary %>%
  select(price, age, gender, urban) %>%
  as.matrix()

#Predict purchase probabilities for the actual data
xgb_pred_prob_insample <- predict(xgb_model, newdata = xgb_insample_matrix)


xgb_pred_binary <- as.integer(xgb_pred_prob_insample > 0.5)
xgb_validation_df <- sim_binary %>%
  mutate(
    predicted_prob     = xgb_pred_prob_insample,
    predicted_purchase = xgb_pred_binary
  ) %>%
  select(id, time, price, age, gender, urban, purchase,
         predicted_prob, predicted_purchase)

# Calculate in-sample hit rate
xgb_hit_rate_insample <- mean(xgb_validation_df$predicted_purchase == xgb_validation_df$purchase)

cat("In-sample Hit Rate for XGBoost:", round(xgb_hit_rate_insample, 3), "\n")

head(xgb_validation_df, 20)

```

**testing data simulation**
```{r}

set.seed(999) 
N_test <- 20
T <- 10
price_range <- c(3, 8)

age_test <- sample(18:70, N_test, replace = TRUE)
gender_test <- sample(c(0, 1), N_test, replace = TRUE)
urban_test <- sample(c(0, 1), N_test, replace = TRUE)

beta_i_test <- mvrnorm(n = N_test, mu = mu_, Sigma = Sigma_)

sim_binary_test <- data.frame()

for (i in 1:N_test) {
  for (t in 1:T) {
    price <- runif(1, price_range[1], price_range[2])
    X <- c(1, price, age_test[i], gender_test[i], urban_test[i])
    utility <- sum(X * beta_i_test[i, ])
    p <- 1 / (1 + exp(-utility))
    y <- rbinom(1, 1, p)
    
    sim_binary_test <- rbind(sim_binary_test, data.frame(
      id = i + N,  # ensure IDs don't overlap with training data
      time = t,
      price = price,
      age = age_test[i],
      gender = gender_test[i],
      urban = urban_test[i],
      purchase = y
    ))
  }
}

head(sim_binary_test)

```

**ground truth for the testing data**
```{r}
purchase_prob <- function(X, beta) {
  plogis(sum(X * beta))
}

expected_revenue <- function(price, X, beta) {
  price * purchase_prob(X, beta)
}
price_options <- c(3, 8)
ground_truth_test <- data.frame()

# Loop through each of the 20 out-of-sample individuals
for (i in 1:N_test) {
  # Extract covariates for test individuals
  indiv_covs <- c(age_test[i], gender_test[i], urban_test[i])

  X3 <- c(1, 3, indiv_covs)
  X8 <- c(1, 8, indiv_covs)
  
  prob_3 <- purchase_prob(X3, beta_i_test[i, ])
  prob_8 <- purchase_prob(X8, beta_i_test[i, ])
  
  rev_3 <- expected_revenue(3, X3, beta_i_test[i, ])
  rev_8 <- expected_revenue(8, X8, beta_i_test[i, ])
  
  opt_price <- ifelse(rev_3 > rev_8, 3, 8)
  
  ground_truth_test <- rbind(
    ground_truth_test,
    data.frame(
      id = i + N,   # IDs start after training data
      price = 3,
      predicted_prob = prob_3,
      predicted_revenue = rev_3,
      optimal_price = opt_price
    ),
    data.frame(
      id = i + N,
      price = 8,
      predicted_prob = prob_8,
      predicted_revenue = rev_8,
      optimal_price = opt_price
    )
  )
}

head(ground_truth_test)

total_optimal_revenue_test <- ground_truth_test %>%
  filter(price == optimal_price) %>%
  summarise(total_revenue = sum(predicted_revenue)) %>%
  pull(total_revenue)

cat("Total ground truth optimal revenue for TEST sample:", total_optimal_revenue_test, "\n")


```

**Fitting the trained XGBoost on testing data, find the pricing policy and the expected total revenue  **
```{r}

# Build testing covariates
testing_covs <- sim_binary_test %>%
  group_by(id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(id, age, gender, urban)

price_given <- c(3, 8)
newdata_test <- testing_covs %>%
  tidyr::crossing(price = price_given) %>%
  arrange(id, price)

# Predict purchase probabilities using xgb_model
xgb_test_matrix <- xgb.DMatrix(data = as.matrix(newdata_test %>% select(price, age, gender, urban)))

xgb_pred_probs <- predict(xgb_model, xgb_test_matrix)

xgb_predictions_test <- newdata_test %>%
  mutate(
    xgb_predicted_prob = xgb_pred_probs,
    xgb_predicted_revenue = xgb_predicted_prob * price
  )

#Select optimal price per individual under XGB policy
xgb_best_prices_test <- xgb_predictions_test %>%
  group_by(id) %>%
  slice_max(xgb_predicted_revenue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(id, xgb_optimal_price = price)

xgb_predictions_test <- xgb_predictions_test %>%
  left_join(xgb_best_prices_test, by = "id")

head(xgb_predictions_test)

# Evaluate XGB pricing policy on testing data
xgb_eval_test <- xgb_best_prices_test %>%
  left_join(testing_covs, by = "id") %>%
  mutate(
    alpha       = beta_i_test[id - N, 1],
    beta_price  = beta_i_test[id - N, 2],
    beta_age    = beta_i_test[id - N, 3],
    beta_gender = beta_i_test[id - N, 4],
    beta_urban  = beta_i_test[id - N, 5]
  )

# Define purchase probability & revenue using ground truth
purchase_prob <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  linpred <- alpha + beta_price * p + beta_age * a + beta_gender * g + beta_urban * u
  plogis(linpred)
}

expected_revenue <- function(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban) {
  p * purchase_prob(p, a, g, u, alpha, beta_price, beta_age, beta_gender, beta_urban)
}

# Compute true expected revenue for chosen optimal prices
xgb_eval_test <- xgb_eval_test %>%
  mutate(
    true_prob = purchase_prob(xgb_optimal_price, age, gender, urban,
                              alpha, beta_price, beta_age, beta_gender, beta_urban),
    true_revenue = expected_revenue(xgb_optimal_price, age, gender, urban,
                                    alpha, beta_price, beta_age, beta_gender, beta_urban)
  )

total_xgb_policy_revenue_test <- sum(xgb_eval_test$true_revenue)

cat("Total expected revenue on testing data using XGB policy:",
    total_xgb_policy_revenue_test, "\n")

```

**frequncy table of two prices under the XGBoost pricing policy on testing dataset**
```{r}
library(dplyr)

xgb_counts2 <- xgb_predictions_test %>%
  group_by(xgb_optimal_price) %>%
  summarise(count = n() / 2) %>%
  arrange(xgb_optimal_price)

print(xgb_counts2)
```

**hit rate on testing data**
```{r}
#Same as what training data did
xgb_test_matrix <- xgb.DMatrix(data = as.matrix(sim_binary_test %>%
                                                 select(price, age, gender, urban)))

xgb_pred_prob_test <- predict(xgb_model, xgb_test_matrix)

xgb_pred_binary_test <- as.integer(xgb_pred_prob_test > 0.5)

xgb_validation_test_df <- sim_binary_test %>%
  mutate(
    predicted_prob = xgb_pred_prob_test,
    predicted_purchase = xgb_pred_binary_test
  ) %>%
  select(id, time, price, age, gender, urban, purchase, predicted_prob, predicted_purchase)

head(xgb_validation_test_df)

xgb_hit_rate_test <- mean(xgb_validation_test_df$predicted_purchase == xgb_validation_test_df$purchase)

cat("Out-of-sample Hit Rate for XGBoost on Testing Data:", xgb_hit_rate_test, "\n")

```
